---
title: "INF2400-Millennials and Finances_v1"
author: "Asel, Rachel, Shabrina, Shweta"
date: "07/07/2020"
output:
  pdf_document: default
  html_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dbplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(stringr)
library(skimr)
library(janitor)
library(cansim) # A library to access Statistics Canada datasets

```


# Financial Security Survey dataset.
```{r, echo = FALSE}

# Downloaded 2016 Survey of Financial Security by Pensions and Wealth Surveys Section, Income Statistics Division
fin_security <- get_cansim("11100016")

fin_security <- janitor::clean_names(fin_security)

```

## Filtering age groups and dollar values
```{r, echo = FALSE}
cleaned_fin_sec_data <- 
  fin_security %>% 
  filter(economic_family_type == "Economic families and persons not in an economic family", uom == "2016 constant dollars", confidence_intervals == "Estimate", statistics == "Total values") %>%
  filter(age_group != "All ages") %>% 
  filter(age_group != "Under 65 years") %>% 
  select(ref_date,
         geo,
         assets_and_debts,
         age_group,
         value) %>% 
  mutate(age = case_when(
    age_group == "Under 35 years" ~ 1,
    age_group == "35 to 44 years" ~ 2,
    age_group == "45 to 54 years" ~ 3,
    age_group == "55 to 64 years" ~ 4,
    age_group == "65 years and older" ~ 5))


```

## Break down the assets and debts column into variables and filter to provinces
```{r, echo = FALSE}

provinces_fin_sec <- cleaned_fin_sec_data %>% 
  pivot_wider(names_from = assets_and_debts, values_from = value) %>% 
  filter(geo != "Canada") %>% 
  filter(geo != "Atlantic") %>% 
  filter(geo != "Québec, Quebec") %>% 
  filter(geo != "Montréal, Quebec") %>% 
  filter(geo != "Toronto, Ontario") %>% 
  filter(geo != "Prairies") %>% 
  filter(geo != "Winnipeg, Manitoba") %>% 
  filter(geo != "Calgary, Alberta") %>% 
  filter(geo != "Edmonton, Alberta") %>% 
  filter(geo != "Vancouver, British Columbia")

provinces_fin_sec <- janitor::clean_names(provinces_fin_sec)

write.csv(provinces_fin_sec, "/Users/aselkushkeyeva/Desktop/provinces_fin_sec.csv")

```

## Plot financial assets non pension and age_group
```{r, echo = FALSE}
provinces_fin_sec %>% 
  ggplot(mapping = aes(x = age_group, y = financial_assets_non_pension/1000)) +
  geom_col() +
  facet_wrap(vars(ref_date)) +
  theme_light() +
  labs(x = "Age groups",
      y = "Financial Assests (non-pension) in Million CAD",
      title = "Financial assets across age groups in Million CAD")

```

## Plot student loans and age_group
```{r, echo = FALSE}
provinces_fin_sec %>% 
  ggplot(mapping = aes(x = age_group, y = student_loans)) +
  geom_col() +
  facet_wrap(vars(ref_date)) +
  theme_light() +
  labs(title = "Student loans")

```

## Plot mortgages and age_group
```{r, echo = FALSE}
provinces_fin_sec %>% 
  ggplot(mapping = aes(x = age_group, y = mortgages/1000)) +
  geom_col() +
  facet_wrap(vars(ref_date)) +
  theme_light() +
  labs(title = "Mortgages")

```


## Plot credit card debt and age_group

Credit card debts ar not as high for millennials as for elder age groups, 45 to 54 years old being the highest in 2016.


```{r, echo = FALSE}
provinces_fin_sec %>% 
  ggplot(mapping = aes(x = age_group, y = credit_card_and_installment_debt)) +
  geom_col() +
  facet_wrap(vars(ref_date)) +
  theme_light() +
  labs(title = "Credit cards")

```

## Plot line of creditt and age_group

```{r, echo = FALSE}
provinces_fin_sec %>% 
  ggplot(mapping = aes(x = age_group, y = line_of_credit)) +
  geom_col() +
  facet_wrap(vars(ref_date)) +
  theme_light() +
  labs(title = "Line of credit")

```

## Plot RRSPs across provinces

```{r, echo = FALSE}
provinces_fin_sec %>% 
  ggplot(mapping = aes(x = geo,
                       y = registered_retirement_savings_plans_rrs_ps_registered_retirement_income_funds_rri_fs_locked_in_retirement_accounts_lir_as_and_other/1000)) +
  geom_col() +
  facet_wrap(vars(ref_date)) +
  theme_light() +
  labs(title = "RRSPs, RRIFs, LIRAs and others") +
  coord_flip()

```

## Extract 1999 and 2016 - Gen X and Millennials - and plot financial assets

```{r, echo = FALSE}

prov_noAtl_1999_2016 <- provinces_fin_sec %>% 
  filter(ref_date != 2005) %>% 
  filter(ref_date != 2012) %>% 
  filter(age_group == "Under 35 years") %>% 
  filter(geo != "Newfoundland and Labrador") %>% 
  filter(geo != "Prince Edward Island") %>% 
  filter(geo != "Nova Scotia") %>% 
  filter(geo != "New Brunswick")
```
```{r, echo = FALSE}
prov_noAtl_1999_2016 %>% 
  ggplot(mapping = aes(x = geo,
                       y = registered_retirement_savings_plans_rrs_ps_registered_retirement_income_funds_rri_fs_locked_in_retirement_accounts_lir_as_and_other)) +
  geom_col() +
  theme_light() +
  labs(title = "Registered retirement funds")+ xlab("Provinces")+
  ylab("Value in Million CAD") +
  facet_wrap(vars(ref_date)) +
  coord_flip()

write.csv(prov_noAtl_1999_2016, "/Users/aselkushkeyeva/Desktop/prov_no_Atlantic_1999_2016.csv")

```
## ONTARIO only

```{r, echo = FALSE}

prov_ontario_1999_2016 <- provinces_fin_sec %>% 
  filter(ref_date != 2005) %>% 
  filter(ref_date != 2012) %>% 
  filter(age_group == "Under 35 years") %>% 
  filter(geo == "Ontario")
  
```
```{r, echo = FALSE}
prov_ontario_1999_2016 %>% 
  ggplot(mapping = aes(x = mutual_funds_investment_funds_and_income_trusts,
                       y = ref_date)) +
  geom_col() +
  theme_light() +
  labs(title = "Ontario Mutual Funds") +
  xlab("Value in Million CAD") + ylab("1999 and 2016") +
  coord_flip()

write.csv(prov_ontario_1999_2016, "/Users/aselkushkeyeva/Desktop/prov_Ontario_1999_2016.csv")

```
## Ontario only - Rrsps

```{r, echo = FALSE}

prov_ontario_1999_2016 <- provinces_fin_sec %>% 
  filter(ref_date != 2005) %>% 
  filter(ref_date != 2012) %>% 
  filter(age_group == "Under 35 years") %>% 
  filter(geo == "Ontario")
  
```
```{r, echo = FALSE}
prov_ontario_1999_2016 %>% 
  ggplot(mapping = aes(x = registered_retirement_savings_plans_rrs_ps_registered_retirement_income_funds_rri_fs_locked_in_retirement_accounts_lir_as_and_other,
                       y = ref_date)) +
  geom_col() +
  theme_light() +
  labs(title = "") +
  xlab("Value in Million CAD") + ylab("1999 and 2016")+
  coord_flip()

#write.csv(prov_ontario_1999_2016, "/Users/aselkushkeyeva/Desktop/prov_Ontario_1999_2016.csv")

```

```{r, echo = FALSE}
prov_noAtl_1999_2016 %>% 
  ggplot(mapping = aes(x = geo,
                       y = financial_assets_non_pension)) +
  geom_col() +
  theme_light() +
  labs(title = "Financial assets (non-pension)") +
  xlab("Provinces") + ylab("Value in Million CAD")+
  facet_wrap(vars(ref_date)) +
  coord_flip()
```
```{r, echo = FALSE}
prov_noAtl_1999_2016 %>% 
  ggplot(mapping = aes(x = geo,
                       y = private_pension_assets)) +
  geom_col() +
  theme_light() +
  labs(title = "Private pension assets") +
  facet_wrap(vars(ref_date)) +
  coord_flip()
```
```{r, echo = FALSE}
# ==========Mutual funds data available only for Ontario
prov_noAtl_1999_2016 %>% 
  ggplot(mapping = aes(x = geo,
                       y = mutual_funds_investment_funds_and_income_trusts)) +
  geom_col() +
  theme_light() +
  labs(title = "Mutual funds") +
  facet_wrap(vars(ref_date)) +
  coord_flip()
```




## Extract 1999 and 2016 - Gen X and Millennials - from the original dataset with assets and debts in one column and plot

```{r, echo = FALSE}

cleaned_1999_2016 <- cleaned_fin_sec_data %>% 
  filter(ref_date != 2005) %>% 
  filter(ref_date != 2012) %>% 
  filter(geo != "Canada") %>% 
  filter(geo != "Atlantic") %>% 
  filter(geo != "Québec, Quebec") %>% 
  filter(geo != "Montréal, Quebec") %>% 
  filter(geo != "Toronto, Ontario") %>% 
  filter(geo != "Prairies") %>% 
  filter(geo != "Winnipeg, Manitoba") %>% 
  filter(geo != "Calgary, Alberta") %>% 
  filter(geo != "Edmonton, Alberta") %>% 
  filter(geo != "Vancouver, British Columbia") %>% 
  filter(age_group == "Under 35 years") %>% 
  pivot_wider(names_from = assets_and_debts, values_from = value)

cleaned_1999_2016 <- janitor::clean_names(cleaned_1999_2016)
```
```{r, echo = FALSE}
cleaned_rows_1999_2016 <- cleaned_1999_2016 %>% 
  rename(RRSPs_RRIFs_LIRAs_other = registered_retirement_savings_plans_rrs_ps_registered_retirement_income_funds_rri_fs_locked_in_retirement_accounts_lir_as_and_other) %>% 
  rename(mutual_funds_income_trusts = mutual_funds_investment_funds_and_income_trusts) %>% 
  rename(TFSAs = tax_free_saving_accounts_tfsa) %>% 
  rename(employer_sponsored_pension_plans = employer_sponsored_registered_pension_plans_ep_ps)
```
```{r, echo = FALSE}
longer_1999_2016 <- cleaned_rows_1999_2016  %>% 
  pivot_longer(cols = c(total_assets,
                        private_pension_assets,
                        RRSPs_RRIFs_LIRAs_other,
                        employer_sponsored_pension_plans,
                        financial_assets_non_pension,
                        deposits_in_financial_institutions,
                        mutual_funds_income_trusts,
                        stocks,
                        bonds_saving_and_other,
                        TFSAs,
                        other_financial_assets,
                        non_financial_assets,
                        principal_residence,
                        other_real_estate,
                        vehicles,
                        other_non_financial_assets,
                        equity_in_business,
                        total_debts,
                        mortgages,
                        mortgage_on_principal_residence,
                        mortgage_on_other_real_estate,
                        line_of_credit,
                        credit_card_and_installment_debt,
                        student_loans,
                        vehicle_loans,
                        other_debt,
                        net_worth_assets_less_debts), 
               names_to = "assets_debts",
               values_to = "value")
```
```{r, echo = FALSE}
longer_1999_2016 %>% 
  ggplot(mapping = aes(x = assets_debts, y = value)) +
  geom_col() +
  theme_light() +
  labs(title = "All assests and debts for Gen X and Millennials") +
  facet_wrap(vars(ref_date)) +
  coord_flip()

write.csv(longer_1999_2016, "/Users/aselkushkeyeva/Desktop/longer_1999_2016.csv")
```

## Extract 2016 data

```{r, echo = FALSE}

provinces_2016 <- provinces_fin_sec %>% 
  filter(ref_date == 2016)
```

## Plot 2016 data
```{r, echo = FALSE}

provinces_2016  %>% 
  ggplot(mapping = aes(x = geo,
                       y = tax_free_saving_accounts_tfsa)) +
  geom_col() +
  theme_light() +
  labs(title = "TFSA Values in 2016") +
  xlab("Provinces") +
  ylab("Tax Free Savings Accounts Value in Million CAD")+
  facet_wrap(vars(age_group)) +
  coord_flip()
 # theme(legend.position="bottom")
```


\newpage
# References

- Ascione, L. Higher-ed leaders: Here’s how Generation Z learns best.
https://www.ecampusnews.com/2019/02/15/higher-ed-leaders-generation-z-learns-best/

- BBC. (2017). The avocado toast index: How many breakfasts to buy a house?
https://www.bbc.com/worklife/article/20170530-the-avocado-toast-index-how-many-breakfasts-to-buy-a-house

- Campbell, D., Hunt, J., & Wissink, R. (2019). The Millennial Report.
http://cdn3.reincanada.com/Research/REIN_The_Millennial_Report_DIGITAL.pdf

- CFA Institute. (2018, October). Uncertain futures: 7 myths about millennials and investing.  https://www.cfainstitute.org/-/media/documents/support/advocacy/1801081-insights-millennials-and-investing-booklet.ashx

- Chen, J. (2020, June 8). Millennials: Finances, Investing, and Retirement. Investopedia.
https://www.investopedia.com/terms/m/millennial.asp

- Cooper, L. (2016, October). Millennials. RBC Economics. Research. http://www.rbc.com/economics/economic-reports/pdf/other-reports/Millennials-October%202016.pdf

- Cooper, P. (2018, January 8). How Degree Inflation Weakens The Economy.
https://www.forbes.com/sites/prestoncooper2/2018/01/08/employers-demanding-college-degrees-weaken-the-economy/#2ae113986b11

- Crowe, Anna. (2016, August 3). Growth Hacking Your Content on Medium. Search Engine Journal.
https://www.searchenginejournal.com/12-step-guide-growth-medium/167771/#:~:text=In%20December%202015%2C%20according%20to,Medium%20has%2061.7%20million%20visits.&text=In%20April%202015%2C%20Native%20Advertising,earn%20six%20figures%20or%20more.

- The Environics Institute. (2016). Canadian Millennials Social Values Survey - 2016.
https://www.environicsinstitute.org/docs/default-source/project-documents/canadian-millennial-social-values-study/canadian-millennials-social-values-study---data-tables.pdf

- The Environics Institute. (2017). Canadian Millennial Social Values Study.
https://www.environicsinstitute.org/projects/project-details/canadian-millennial-social-values-study

- Franklin, H. (2019, February 14). Why repaying student debt early is the best investment you can make.
https://www.moneysense.ca/columns/ask-moneysense/repay-student-debt-before-investing/

- Firke, S. (2020). janitor: Simple Tools for Examining and Cleaning Dirty Data. R
  package version 2.0.1. https://CRAN.R-project.org/package=janitor

- Gaviola, A. (2020, May 19). Canada, Here's What You Need to Do About Your Student Loans Right Now.
https://www.vice.com/en_ca/article/z3evey/canada-heres-what-you-need-to-do-about-your-student-loans-right-now

- Goldman Sachs. (2020). Millennials. Coming of age. https://www.goldmansachs.com/insights/archive/millennials/

- Heisz, A. & Richards, E. (2019, April 18). Economic well-being across generations of young Canadians: Are millennials better or worse off? https://www150.statcan.gc.ca/n1/pub/11-626-x/11-626-x2019006-eng.htm

- Iacurci, G. (2020, February 4). The gig economy has ballooned by 6 million people since 2010. Financial worries may follow.
https://www.cnbc.com/2020/02/04/gig-economy-grows-15percent-over-past-decade-adp-report.html

- Kim, L. (2018, November 13). 10 Insanely Good Reasons You Should Publish On Medium. The WordStream Blog.
https://www.wordstream.com/blog/ws/2016/07/14/medium-publishing

- Knueven, L. (2020, February 4). The 7 best ways to build wealth starting today, according to financial planners.
https://www.businessinsider.com/personal-finance/best-ways-to-build-wealth-starting-today-2019-8

- KPMG. (2019, December 5). Owning a home is becoming a pipedream for many millennials.
https://home.kpmg/ca/en/home/media/press-releases/2019/12/owning-a-home-is-a-pipedream-for-many-millennials.html

- Lalonde, D. (2018, November 30). Millennials save but why don’t they invest? https://abacusdata.ca/millennials-save-but-why-dont-they-invest/

- Lemke, T. (2020, April 2). Millennials investing habits you can learn from in 2019. https://www.thebalance.com/how-millennials-are-investing-these-days-4174110

- Levin, S. (2017, May 15). Millionaire tells millennials: if you want a house, stop buying avocado toast.
https://www.theguardian.com/lifeandstyle/2017/may/15/australian-millionaire-millennials-avocado-toast-house

- Martin, J. (2016, September). The Game: Easier or Tougher?
https://pepso.ca/documents/millennial-research-summary-v7-rev-1.pdf

- Ratiu, G. (2019, December 4). 2020 Housing Market Predictions.
https://www.realtor.com/research/2020-national-housing-forecast/

- RBC Economics.(2016, October). Millennials.
http://www.rbc.com/economics/economic-reports/pdf/other-reports/Millennials-October%202016.pdf

- Reddit. (2015). Why millennials don't invest in the stock market.
https://www.reddit.com/r/investing/comments/327z2j/why_millennials_dont_invest_in_the_stock_market/

- RStudio Team. (2020). RStudio: Integrated Development for R. RStudio, PBC, Boston, MA URL.         http://www.rstudio.com/

- Sheppard, C. (2018, September 6). Adult Shamed For Living With His Parents Has Epic Clapback.
https://www.huffpost.com/entry/adult-shamed-for-living-with-his-parents-has-epic-clapback_n_5b9168d4e4b0511db3e04741

- Shkolnik, D. (2020). cansim: Accessing Statistics Canada Data Table and Vectors. R
  package version 0.3.5. https://CRAN.R-project.org/package=cansim
  
- Statistics Canada. (2016). The Survey of Financial Security.                      https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1110001601

- Skelton, Z. L. (2016, November 25). The Phone Habits of Millennials – and How to Market to Them. B2B News. https://www.b2bnn.com/2016/11/phone-habits-millennials-market/

- Tencer, D. (2016, September 7). Canadian Tuitions 45% Higher Than A Decade Ago, And Still Rising. 
https://www.huffingtonpost.ca/2016/09/07/tuition-rising-canada_n_11894908.html?guccounter=1

- Waring, E., Quinn, M., McNamara, A., Arino de la Rubia, E., Zhu, H. & Ellis, S. (2020). 
  skimr: Compact and Flexible Summaries of Data. R package version
  2.1.2. https://CRAN.R-project.org/package=skimr

- Wickham, H. (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York.

- Wickham, H. (2019). stringr: Simple, Consistent Wrappers for Common String Operations. 
  R package version 1.4.0. https://CRAN.R-project.org/package=stringr

- Wickham, H. & Henry, L. (2020). tidyr: Tidy Messy Data. R package version
  1.1.0. https://CRAN.R-project.org/package=tidyr
  
- Wickham et al. (2019). Welcome to the tidyverse. Journal of Open Source Software,
  4(43), 1686. https://doi.org/10.21105/joss.01686
  
- Williams, C. (2003, November). Finances in the golden years.
https://www150.statcan.gc.ca/n1/pub/75-001-x/01103/6676-eng.html

- Young, A., Wolfe, R., & Sellery, B. (2018). Getting Started: Human-Centred Solutions to Engage Ontario Millennials in Investing
https://www.osc.gov.on.ca/documents/en/Investors/inv_research_20180712_getting-started.pdf
